---
title: "W12-L3-Examples"
author: "Dr. Coleman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in packages

```{r message=FALSE}
library(tidyverse)
library(tidymodels)
library(lubridate)
library(kknn)
library(ranger)
```

# Example: mtcars

Train/test split

```{r}

# Perform the split
dat_split <- initial_split(mtcars)
dat_split

# Training dataset
dat_train <- training(dat_split)
dat_train

# Testing dataset (save for later)
dat_test <- testing(dat_split)
dat_test

# Visualize training data
dat_train |>
    ggplot(aes(hp, mpg)) +
    geom_point()

```

## Recipe, model, workflow

Start with linear regression

```{r}

# Define recipe
rec <- dat_train |>
    recipe(mpg ~ cyl + disp + hp + wt)
rec

# Define model
model <- linear_reg() |>
    set_engine("lm") |>
    set_mode("regression")
model

# Define workflow
wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)
wf
```

## Cross-validation

Check performance using cross-validation

```{r}

# Randomly split the data into folds
folds <- vfold_cv(dat_train, v = 10)
folds

# Fit models using resampling
res <- wf |>
    fit_resamples(resamples = folds)

# Collect metrics
res |>
    collect_metrics()

# Fit model without cross validation
fit <- wf |>
    fit(data = dat_train)

# Plot predicted and actual outcome (mpg)
fit |>
    extract_fit_parsnip() |>
    augment(dat_train) |>
    ggplot(aes(.pred, mpg)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)
```

## Different models

At this point we could try different models for our data. These could include k-NN or random forest. This could also involve preprocessing changes. The recipe, model performance, and workflow may be updated.

## Settled on a model

Now, do final estimate of performance

```{r}

# Fit the final best model to the training set and evaluate the test set
final <- wf |>
    last_fit(split = dat_split)

# Collect metrics
final |>
    collect_metrics()
```

## Plot observed vs. predicted

```{r}

# Note: we are using the test dataset
final |>
    extract_fit_parsnip() |>
    augment(dat_test) |>
    ggplot(aes(.pred, mpg)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```

# Next example: Particulate matter

```{r}
dat0 <- read_csv("planet.csv.gz")
dat0
```

## Show one image

```{r}
dat0 |>
    slice(500) |>   ## Show the 1st image (change this to see others)
    select(-pmFRM) |>  ## Remove the label column (don't need it for now)
    pivot_longer(everything()) |> ## Pivot to long form
    bind_cols(expand_grid(x = 1:64, y = 1:64)) |>  ## Combined row/column numbers
    ggplot(aes(x = y, y = x)) +   ## Plot it so that it's right side up
    geom_raster(aes(fill = value)) +
    labs(x = NULL, y = NULL) +
    scale_fill_gradient(low = "white", high = "black") +
    coord_fixed() +
    theme_bw() +
    scale_y_reverse() +
    theme(legend.position = "none")
```

## Subset the variables to save time

```{r}

set.seed(2026-04-10)
var_use <- sample(4096, 64)
dat <- dat0 |>
    select(pmFRM, num_range(prefix = "pixel",
                            range = var_use))
dat

```

## Show which pixels were selected

```{r}
expand_grid(x = 1:64, y = 1:64) |>
    mutate(z = NA) |>
    mutate(z = replace(z, var_use, 1)) |>
    ggplot(aes(y, x)) +
    geom_raster(aes(fill = z)) +
    scale_y_reverse() +
    coord_fixed()
```

## Immediately split the data into training and test sets

```{r}
dat_split <- initial_split(dat)
dat_split

dat_train <- training(dat_split)
dat_train

pixel_col <- colnames(dat_train) |> # colnames of df
  str_subset('pixel[0-9]*') |> # get those with pixel in them
  sample(1) # sample one randomly

# Visualize this
dat_train |>
    rename(rand_pixel =  all_of(pixel_col)) |> # rename column
    ggplot(aes(rand_pixel, pmFRM)) +
    geom_point() +
    geom_smooth()
```

Show pmFRM with all pixels

```{r}

dat_train |>
    pivot_longer(-pmFRM) |>
    ggplot(aes(value, pmFRM)) +
    geom_point(alpha = 1/10) +
    geom_smooth() +
    facet_wrap(vars(name))
```

## Start with linear regression

```{r}

rec <- dat_train |>
    recipe(pmFRM ~ .) |> # . means all others
    step_normalize(starts_with("pixel"))
rec

model <- linear_reg() |>
    set_engine("lm") |>
    set_mode("regression")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)
wf

res <- fit(wf, data = dat_train)
```

## Check performance on the complete training data

```{r}

res |>
    extract_fit_engine() |>
    summary()

```

## Check performance using cross-validation

```{r}

folds <- vfold_cv(dat_train, v = 10)
folds
res <- fit_resamples(wf, resamples = folds)
res |>
    collect_metrics()

```

## Try k-NN model

```{r}

rec <- dat_train |>
    recipe(pmFRM ~ .) |>
    step_normalize(starts_with("pixel"))

model <- nearest_neighbor(neighbors = 10) |>
    set_engine("kknn") |>
    set_mode("regression")

wf <- workflow() |>
    add_model(model) |>
    add_recipe(rec)
wf

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds)

res |>
    collect_metrics()
```

## Tune for the optimal number of neighbors

```{r}

# Define model and note k as a tuning parameter
model <- nearest_neighbor(neighbors = tune("k")) |>
    set_engine("kknn") |>
    set_mode("regression")

# Define workflow
wf <- workflow() |>
    add_model(model) |>
    add_recipe(rec)
wf

# Cross validation split
folds <- vfold_cv(dat_train, v = 10)

# Find optimal k
res <- tune_grid(wf, resamples = folds,
                 grid = tibble(k = c(10, 12, 13, 15, 
                                     17, 18, 20, 25)))

# Get metrics
res |>
    collect_metrics()

# Visualize with ggplot
res |>
    collect_metrics() |>
    filter(.metric == "rmse") |>
    ggplot(aes(k, mean)) +
    geom_point() +
    geom_line()
```

```{r}

# Find best metrics

res |>
    show_best(metric = "rmse")

res |>
    show_best(metric = "rsq")

res |>
    collect_metrics() |>
    filter(.metric == "rsq") |>
    ggplot(aes(k, mean)) +
    geom_point() +
    geom_line()

```

## Try k-NN with PCA on predictors

```{r}

rec <- dat_train |>
    recipe(pmFRM ~ .) |>
    step_normalize(starts_with("pixel")) |>
    step_pca(starts_with("pixel"), num_comp = 3)


model <- nearest_neighbor(neighbors = 17) |>
    set_engine("kknn") |>
    set_mode("regression")

wf <- workflow() |>
    add_model(model) |>
    add_recipe(rec)

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds)
res |>
    collect_metrics()

```

## Tune PCA Components

```{r}

rec <- dat_train |>
    recipe(pmFRM ~ .) |>
    step_normalize(starts_with("pixel")) |>
    step_pca(starts_with("pixel"), num_comp = tune("pca"))

model <- nearest_neighbor(neighbors = 15) |>
    set_engine("kknn") |>
    set_mode("regression")
wf <- workflow() |>
    add_model(model) |>
    add_recipe(rec)


res <- tune_grid(wf, resamples = folds,
                 grid = tibble(pca = c(1, 5, 10, 15, 20)))

res |>
    show_best(metric = "rmse")

res |>
    show_best(metric = "rsq")

```

## Try random forest instead

```{r}

rec <- dat_train |>
    recipe(pmFRM ~ .) |>
    step_normalize(starts_with("pixel"))

model <- rand_forest(mtry = 5) |>
    set_engine("ranger") |>
    set_mode("regression")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)
wf

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds)

res |>
    collect_metrics()

```

## Try a grid of tuning parameters

```{r}

model <- rand_forest(mtry = tune("mtry"),
                     min_n = tune("min_n")) |>
    set_engine("ranger") |>
    set_mode("regression")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

```

## Fit model over grid of tuning parameters

```{r}

res <- tune_grid(wf, resamples = folds,
                 grid = expand.grid(mtry = c(1, 2, 5),
                                    min_n = c(3, 5)))
res |>
    collect_metrics()

res |>
    show_best(metric = "rmse")

res |>
    show_best(metric = "rsq")

res |>
    select_best(metric = "rmse")

```

## Fit the best model obtained from tuning

```{r}

model <- rand_forest(mtry = 2,
                     min_n = 3) |>
    set_engine("ranger") |>
    set_mode("regression")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

```

## Fit final model to entire training set; evaluate on test set

```{r}

final <- wf |>
    last_fit(split = dat_split)

final |>
    collect_metrics()

```

## Plot the observed PM2.5 values vs. model predictions

```{r}

final |>
    collect_predictions() |>
    ggplot(aes(.pred, pmFRM)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```

# Final example: Spotify and YouTube

Dataset from Kaggle: <https://www.kaggle.com/datasets/salvatorerastelli/spotify-and-youtube>

```{r}

dat0 <- read_csv("Spotify_Youtube.csv.gz")
dat0
dat0 |>
    glimpse()

dat <- dat0 |>
    select(-1) |>
    select(where(is.numeric)) |>
    filter(if_all(everything(), ~ !is.na(.x))) |> ## Remove rows w/NA values
    sample_frac(0.1)

dat |>
    glimpse()

```

## Histogram of streams

```{r}

dat |>
    ggplot(aes(Stream)) +
    geom_histogram(bins = 10) +
    labs(x = "Streams")

```

## Summarize the distribution of the 'Streams' variable

```{r}

summary(dat$Stream)

```

## Dichotomize Stream variable

```{r}

dat <- dat |>
    mutate(success = ifelse(Stream > 50000000, 1, 0),
           success_f = ifelse(success == 1, "hit", "dud"))

dat |>
    glimpse()

dat_split <- initial_split(dat)
dat_split
dat_train <- training(dat_split)

```

## Look at Liveness

```{r}

dat_train |>
    ggplot(aes(Liveness, success)) +
    geom_point(alpha = 1/20) +
    geom_smooth()

```

## Look at Danceability

```{r}

dat_train |>
    ggplot(aes(Danceability, success)) +
    geom_point(alpha = 1/20) +
    geom_smooth()

dat |>
    glimpse()

```

## Plot a bunch of variables

```{r}

dat_train |>
    select(success, Danceability:Duration_ms) |>
    pivot_longer(-success, names_to = "name", 
                 values_to = "metric") |>
    ggplot(aes(metric, success)) +
    geom_point(alpha = 1/2) +
    geom_smooth(se = FALSE) +
    facet_wrap(vars(name), scales = "free_x")

```

## Look at Duration

```{r}

summary(dat_train$Duration_ms)

```

## Start modeling

```{r}

rec <- dat_train |>
    recipe(success_f ~ Danceability)

rec |>
    prep() |>
    bake(new_data = NULL)

model <- logistic_reg() |>
    set_mode("classification") |>
    set_engine("glm")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

res <- fit(wf, data = dat_train)

res |>
    extract_fit_parsnip() |>
    augment(new_data = dat_train)

tidy(res)

```

## Get regular glm() summary

```{r}

res |>
    extract_fit_engine() |>
    summary()

```

## Create 10-folds for cross validation

```{r}

folds <- vfold_cv(dat_train, v = 10)
folds

res <- fit_resamples(wf, resamples = folds)
res |>
    collect_metrics()

```

## Try adding more variables

```{r}

rec <- dat_train |>
    recipe(success_f ~ Danceability + Acousticness + Energy + Loudness
           + Instrumentalness + Speechiness)

rec |>
    prep() |>
    bake(new_data = NULL)

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds)
res |>
    collect_metrics()

```

## Try some other metrics of performance

```{r}

res <- fit_resamples(wf, resamples = folds,
                     metrics = metric_set(accuracy, roc_auc,
                                          ppv, npv, sens, spec))
res |>
    collect_metrics()

```

## Try k-NN instead

```{r}

rec <- dat_train |>
    recipe(success_f ~ Danceability + Acousticness + Energy + Loudness
           + Instrumentalness + Speechiness) |>
    step_normalize(-success_f)

model <- nearest_neighbor(neighbors = 2) |>
    set_engine("kknn") |>
    set_mode("classification")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds,
                     metrics = metric_set(accuracy, roc_auc,
                                          ppv, npv, sens, spec))
res |>
    collect_metrics()

```

## Tune the 'neighbors' parameter

```{r}

model <- nearest_neighbor(neighbors = tune("k")) |>
    set_engine("kknn") |>
    set_mode("classification")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

folds <- vfold_cv(dat_train, v = 10)

res <- tune_grid(wf, resamples = folds,
                 grid = tibble(k = seq(2, 30, 2)))
res |>
    collect_metrics()

```

## Plot performance metrics

```{r}

res |>
    collect_metrics() |>
    filter(.metric == "roc_auc") |>
    ggplot(aes(k, mean)) +
    geom_point() +
    geom_line()

res |>
    show_best(metric = "roc_auc")

```

## Try random forest

```{r}

rec <- dat_train |>
    recipe(success_f ~ Danceability + Acousticness + Energy + Loudness
           + Instrumentalness + Speechiness) |>
    step_normalize(-success_f)

model <- rand_forest() |>
    set_engine("ranger") |>
    set_mode("classification")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds,
                     metrics = metric_set(accuracy, roc_auc,
                                          ppv, npv, sens, spec))
res |>
    collect_metrics()

```

## More variables

```{r}

rec <- dat_train |>
    recipe(success_f ~ Danceability + Acousticness + Energy + Loudness
           + Instrumentalness + Speechiness + Liveness + Duration_ms
           + Valence + Tempo + Key) |>
    step_normalize(-success_f)

model <- rand_forest() |>
    set_engine("ranger") |>
    set_mode("classification")

wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

folds <- vfold_cv(dat_train, v = 10)

res <- fit_resamples(wf, resamples = folds,
                     metrics = metric_set(accuracy, roc_auc,
                                          ppv, npv, sens, spec))
res |>
    collect_metrics()

```

## Tuning for min_n

```{r}

model <- rand_forest(min_n = tune("min_n")) |>
    set_engine("ranger") |>
    set_mode("classification")
wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)
wf
folds <- vfold_cv(dat_train, v = 5)
res <- tune_grid(wf, resamples = folds,
                 grid = tibble(min_n = c(2, 3, 5, 10)))
res |>
    collect_metrics()

res |>
    show_best(metric = "roc_auc")

```

## Remove a variable (Tempo)

```{r}

rec <- dat_train |>
    recipe(success_f ~ Liveness + Loudness + Speechiness +
               Duration_ms + Valence + Instrumentalness + Acousticness +
               Energy) |>
    step_normalize(-success_f)
model <- rand_forest(min_n = 10) |>
    set_engine("ranger") |>
    set_mode("classification")
wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)
folds <- vfold_cv(dat_train, v = 5)
res <- fit_resamples(wf, resamples = folds)
res |>
    collect_metrics()

```

## Final model

```{r}

rec <- dat_train |>
    recipe(success_f ~ Liveness + Tempo + Loudness + Speechiness +
               Duration_ms + Valence + Instrumentalness + Acousticness +
               Energy) |>
    step_normalize(-success_f)
model <- rand_forest(min_n = 10) |>
    set_engine("ranger") |>
    set_mode("classification")
wf <- workflow() |>
    add_recipe(rec) |>
    add_model(model)

```

## Plot the observed hits vs. model predictions

```{r}

res <- last_fit(wf, split = dat_split)
res |>
    collect_metrics()

dat_test <- rec |>
    prep() |>
    bake(new_data = testing(dat_split))

res |>
    extract_fit_parsnip() |>
    augment(new_data = dat_test) |>
    select(success_f, .pred_hit) |>
    ggplot(aes(.pred_hit, success_f)) +
    geom_point(alpha = 1/10)

```

## ROC plot

```{r}

library(plotROC)
res |>
    extract_fit_parsnip() |>
    augment(new_data = dat_test) |>
    select(success_f, .pred_hit) |>
    ggplot(aes(d = success_f,
               m = .pred_hit)) +
    geom_roc()

```
